Scalable statistical genetic data analysis in Python

### 2024-01-08 paper.tex
Move more text to the graveyard

%%% Should only be set by an editor
\jvolume{00}
\jnumber{0}
\jyear{2023}

@@

\section{Background}
% This current phrasing would annoy people, just spewing this
% out to roughly say what I think is relevant, we can refine later.
The study of genetics is currently transforming into a true big-data science.
Since the Human Genome Project, genomics has required working with
signicant volumes of data and the specialised methods
required to transform the raw biological signals into analysable data,
such as assembly[cite], mapping[cite] and variant calling[cite] are compute intensive.
However, beyond these initial steps required to generate the data, the actual
analysis of the genetic variation could typically be performed on a single
computer, and often with ad-hoc software written using shell utilities
and scripting languages such as Perl.
From the Human Genome Project, through 1000 Genomes, and now with
population scale genomics datasets such as UK Biobank, GeL, etc,
the requirements for analysing the data have far outgrown these
methods. Simple queries can require minutes to complete, which
has a negative effect on researcher productivity, as well as limiting
the analyses that can be performed. The financial cost of working
with such datasets is also substantial, which limits accessibility
to an elite set of well-funded groups.

@@

\section{Older text}

@@

@@

@@

@@

\subsubsection{API overview}

@@

\subsubsection{Example}

@@

\subsection{Enabling technologies}

>>>>

%%% Should only be set by an editor
\jvolume{00}
\jnumber{0}
\jyear{2024}

@@

\section{Background}
% This current phrasing would annoy people, just spewing this
% out to roughly say what I think is relevant, we can refine later.

The study of genetics is currently transforming into a true big-data science.
Since the Human Genome Project, genomics has required working with
signicant volumes of data and the specialised methods
required to transform the raw biological signals into analysable data,
such as assembly[cite], mapping[cite] and variant calling[cite] are compute intensive.
However, beyond these initial steps required to generate the data, the actual
analysis of the genetic variation could typically be performed on a single
computer, and often with ad-hoc software written using shell utilities
and scripting languages such as Perl.
From the Human Genome Project, through 1000 Genomes, and now with
population scale genomics datasets such as UK Biobank, GeL, etc,
the requirements for analysing the data have far outgrown these
methods. Simple queries can require minutes to complete, which
has a negative effect on researcher productivity, as well as limiting
the analyses that can be performed. The financial cost of working
with such datasets is also substantial, which limits accessibility
to an elite set of well-funded groups.

@@

\subsection{Storing genetic variation data}

@@

High level discussion of
strategies used to store genotype matrix and accompanying metadata.
Emphasis on access by variant, not sample. Introduce Zarr, and
refer to Fig~\ref{fig-data-storage}A to explain columnar storage,
and Fig~\ref{fig-data-storage}B to show that is competitive
with highly specialised approaches.
Note that the FC simulated dataset is probably a best-case
scenario for methods specialised to patterns in the genotype matrix.
Discuss storing additional data, and advantages of the
columnar approach (refs to genomics tools that do the same).

\subsection{Computing on the genotype matrix}

Compressing data isn't ultimately the point - we want to use it.
We discuss the problem of running computations on the entire
genotype matrix, and how different formats limit what can be done.
We illustrate the discussion by the afdist operation from
the bcftools plugins package. This is perhaps a contrived example, but
we chose it because it requires examining the entire genotype matrix
and there is an efficient implementation available in bcftools.
It is also a good example because the standard approach of
divvying up the genome into chunks via HPC schedulers or workflow
managers is quite tricky.
We discuss the software development aspect, and how we developed
the afdist-savvy program. We discuss the inherent limitations
of the decode-to-vcf strategy, and how this inevitably
leads to bottlenecks. We discuss parallelism, and why
bcftools and genozip can make limited use of available cores.
Also that single-machine parallelism is inevitably constrained
by IO (which is why savvy's thread utilisation drops off).
Figure: \ref{fig-whole-matrix-compute}. Also
supplementary figure ~\ref{fig-whole-matrix-compute-supplemental}.

@@

\subsection{Subsetting the genotype matrix}
Existing work overwhelmingly emphasises accessibility by variant.
We demonstrate that sgkit/zarr is good for getting subsets
in terms of both variants and samples via afdist on various subsets.
Discuss why storing AF values for crude population groupings is
intrinsically bad, and that it should be easy to do calculations
on aribitrary subsets. Also participants regularly drop-out of
large studies like UKB - do you update the AF values each time?

@@

\subsection{Distributed computation}
Why do we need distributed compute? Summary of current
approaches (chunk up along the genome using HPC or workflow
managers). Advantages of true distributed compute, MapReduce
etc. Summary of earlier efforts like ADAM.
Discuss Hail. It's awesome, has enabled current
generation of mega-scale projects. Drawbacks are the
architecture has emphasised performance etc over adaptability
and simplicity. File format is undocumented, etc.
Discuss why the sgkit architecture is suited to distributed
computation. Discuss Zarr's use in cloud stores, and existing
deployments. Discuss Dask.
We do not attempt a head-to-head performance comparison with
Hail here, but defer to Liangde's thesis~\citep{li2022efficient}
with brief summary.
Forward ref to following sections on applications to demonstate
sgkit in a distributed context.

\subsection{Quality control of large datasets}
Outline the importance of QC, and existing approaches.
Summarise sgkit's general QC capabilities.
We demonstrate sgkit on quality control operations on a large
real dataset. We convert large messy VCF with lots of additional
fields. Quote file size numbers. We do some QC operations on
the dataset, using a good size distributed cluster, using
some of the additional fields. Quote times. Discuss Xarray
here, and possibly illustrate the labelled dimensioned arrays
here via a figure (perhaps a screenshot/pdf conversion of
a Jupyter notebook rendering of the big messy dataset,
showing the various fields).

@@

@@

@@

\section*{Text graveyard}

Here's a bunch of text that contains bits that might be useful and readapted.

\section{Enabling technologies}
